---
title: "STAT 626 Project - Version 1"
author: "Ken Marciel"
date: "6/12/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

## US Meat Production

Packages
```{r packages, message = F, warning = F}
library(readxl) # for read_excel function
library(astsa) # for tsplot function
library(dplyr) # data cleaning
```

Data
```{r data, message = F, warning = F}
# read data from Excel file
setwd("C:/Users/keoka/OneDrive - Texas A&M University/Courses/STAT_626/Project/Analysis/Ken/")
meat <- read_excel("MeatStatsFull.xlsx", sheet = 2)
head(meat)

# total monthly red meat and poultry production from January 1983 to April 2021
meat <- meat[464:5, c(1,17)] # changes data from reverse to chronological order
colnames(meat) <- c("Month","MeatProd")
head(meat)
dim(meat) # 460 months = 38 1/3 years
class(meat) # data frame
meat = ts(meat) # change data frame to time series
class(meat) # time series
```

### Exploratory Data Analysis

Upward trend shows that the raw data is not stationary.

```{r analysis}
# 3-term moving average
meat_ma = stats::filter(meat[,2], sides = 2, filter = rep(1/3,3))
par(mfrow = c(2,1))
tsplot(meat[,2], col=2)
lines(meat_ma, col=4)
```

### Forecasting

#### No Transformation

AR models seem to work best, but the data is nonstationary.

```{r}
sarima.for(meat[,2], n.ahead=24, p=1, d=0, q=0) # AR(1)
sarima.for(meat[,2], n.ahead=24, p=2, d=0, q=0) # AR(2)
sarima.for(meat[,2], n.ahead=24, p=3, d=0, q=0) # AR(3)

sarima.for(meat[,2], n.ahead=24, p=0, d=0, q=1) # MA(1)
sarima.for(meat[,2], n.ahead=24, p=0, d=0, q=2) # MA(2)
sarima.for(meat[,2], n.ahead=24, p=0, d=0, q=3) # MA(3)

sarima.for(meat[,2], n.ahead=24, p=1, d=0, q=1) # ARMA(1,1)
sarima.for(meat[,2], n.ahead=24, p=1, d=1, q=1) # ARIMA(1,1,1)
sarima.for(meat[,2], n.ahead=24, p=1, d=2, q=1) # ARIMA(1,2,1)
```

#### Log Transformation

The log transformation removes the trend, resulting in stationarity. Again, the AR models seem to perform best.

```{r}
sarima.for(log(meat[,2]), n.ahead=24, p=1, d=0, q=0) # AR(1)
sarima.for(log(meat[,2]), n.ahead=24, p=2, d=0, q=0) # AR(2)
sarima.for(log(meat[,2]), n.ahead=24, p=3, d=0, q=0) # AR(3)

sarima.for(log(meat[,2]), n.ahead=24, p=0, d=0, q=1) # MA(1)
sarima.for(log(meat[,2]), n.ahead=24, p=0, d=0, q=2) # MA(2)
sarima.for(log(meat[,2]), n.ahead=24, p=0, d=0, q=3) # MA(3)

sarima.for(log(meat[,2]), n.ahead=24, p=1, d=0, q=1) # ARMA(1,1)
sarima.for(log(meat[,2]), n.ahead=24, p=1, d=1, q=1) # ARIMA(1,1,1)
sarima.for(log(meat[,2]), n.ahead=24, p=1, d=2, q=1) # ARIMA(1,2,1)
```

#### First Order Differencing

First order differencing doesn't remove the trend as well as the log transformation.

```{r}
sarima.for(meat[,2], n.ahead=24, p=1, d=1, q=0) # AR(1)
sarima.for(meat[,2], n.ahead=24, p=2, d=1, q=0) # AR(2)
sarima.for(meat[,2], n.ahead=24, p=3, d=1, q=0) # AR(3)

sarima.for(meat[,2], n.ahead=24, p=0, d=1, q=1) # MA(1)
sarima.for(meat[,2], n.ahead=24, p=0, d=1, q=2) # MA(2)
sarima.for(meat[,2], n.ahead=24, p=0, d=1, q=3) # MA(3)

sarima.for(meat[,2], n.ahead=24, p=1, d=1, q=1) # ARIMA(1,1,1)
```

#### Second Order Differencing

Second order differencing creates too much stationarity for AR models, and not enough for MA models.

```{r}
sarima.for(meat[,2], n.ahead=24, p=1, d=2, q=0) # AR(1)
sarima.for(meat[,2], n.ahead=24, p=2, d=2, q=0) # AR(2)
sarima.for(meat[,2], n.ahead=24, p=3, d=2, q=0) # AR(3)

sarima.for(meat[,2], n.ahead=24, p=0, d=2, q=1) # MA(1)
sarima.for(meat[,2], n.ahead=24, p=0, d=2, q=2) # MA(2)
sarima.for(meat[,2], n.ahead=24, p=0, d=2, q=3) # MA(3)

sarima.for(meat[,2], n.ahead=24, p=1, d=2, q=1) # ARIMA(1,1,1)
```


#### Differencing for AR(3)

Differencing doesn't appear to improve the stationarity or forecast. AR(3) on log transformed data seems to work the best of the models investigated so far.

```{r}
sarima.for(log(meat[,2]), n.ahead=24, p=3, d=0, q=0) # AR(1)
sarima.for(log(meat[,2]), n.ahead=24, p=2, d=1, q=0) # AR(2)
sarima.for(log(meat[,2]), n.ahead=24, p=3, d=2, q=0) # AR(3)
```


#### Forecasting using data sets provided by Jocelyn Hunyadi

Jocelyn's code for data loading and cleaning

```{r}
# Load Data
FI_long <- readRDS("FI_long.rds") %>% group_by(Time) %>% 
  summarize(Production_capita = sum(Production_capita, na.rm = TRUE))
FI_long_rmp <- readRDS("FI_long_rmp.rds")
FI_long_big <- readRDS("FI_long_big.rds")
FI_long_small <- readRDS("FI_long_small.rds")

# Time Series Objects:
FI_long_rm <- FI_long_rmp %>% 
  dplyr::select(-Time) %>% arrange(Time2, decreasing = TRUE) %>% 
  filter(Meat_Type == "TotalRedMeat") 
all_rmp <-ts(FI_long_rm$Production, start=c(1983,1), end=c(2021,4), deltat=1/12)

FI_long_p <- FI_long_rmp %>% filter(Meat_Type != "TotalRedMeat") %>% 
  arrange(Time2, decreasing = TRUE)
all_p <- ts(FI_long_p$Production, start=c(1983,1), end=c(2021,4), deltat=1/12)
```


24-Month Forecast: All Red Meat and Poultry

```{r}
sarima.for(diff(log(all_rmp)), n.ahead=24, p=1, d=0, q=0) # AR(1)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=2, d=0, q=0) # AR(2)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=3, d=0, q=0) # AR(3)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=4, d=0, q=0) # AR(4)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=5, d=0, q=0) # AR(5)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=6, d=0, q=0) # AR(6)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=7, d=0, q=0) # AR(7)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=8, d=0, q=0) # AR(8)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=9, d=0, q=0) # AR(9)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=10, d=0, q=0) # AR(10)
sarima.for(diff(log(all_rmp)), n.ahead=24, p=11, d=0, q=0) # AR(11)
sarima.for(diff(log(all_rmp), ), n.ahead=24, p=12, d=0, q=0) # AR(12)
```


Checking the Model Conditions: All Red Meat and Poultry

```{r}
sarima(diff(log(all_rmp)), p=1, d=0, q=0)# AR(1)
sarima(diff(log(all_rmp)), p=2, d=0, q=0) # AR(2)
sarima(diff(log(all_rmp)), p=3, d=0, q=0) # AR(3)
sarima(diff(log(all_rmp)), p=4, d=0, q=0) # AR(4)
sarima(diff(log(all_rmp)), p=5, d=0, q=0) # AR(5)
sarima(diff(log(all_rmp)), p=6, d=0, q=0) # AR(6)
sarima(diff(log(all_rmp)), p=7, d=0, q=0) # AR(7)
sarima(diff(log(all_rmp)), p=8, d=0, q=0) # AR(8)
sarima(diff(log(all_rmp)), p=9, d=0, q=0) # AR(9)
sarima(diff(log(all_rmp)), p=10, d=0, q=0) # AR(10)
sarima(diff(log(all_rmp)), p=11, d=0, q=0) # AR(11)
sarima(diff(log(all_rmp)), p=12, d=0, q=0) # AR(12)
```

Autocorrelation displayed by ACF of Residuals is lowest for AR(12).


24-Month Forecast: All Poultry

```{r}
sarima.for(diff(log(all_p)), n.ahead=24, p=1, d=0, q=0) # AR(1)
sarima.for(diff(log(all_p)), n.ahead=24, p=2, d=0, q=0) # AR(2)
sarima.for(diff(log(all_p)), n.ahead=24, p=3, d=0, q=0) # AR(3)
sarima.for(diff(log(all_p)), n.ahead=24, p=4, d=0, q=0) # AR(4)
sarima.for(diff(log(all_p)), n.ahead=24, p=5, d=0, q=0) # AR(5)
sarima.for(diff(log(all_p)), n.ahead=24, p=6, d=0, q=0) # AR(6)
sarima.for(diff(log(all_p)), n.ahead=24, p=7, d=0, q=0) # AR(7)
sarima.for(diff(log(all_p)), n.ahead=24, p=8, d=0, q=0) # AR(8)
sarima.for(diff(log(all_p)), n.ahead=24, p=9, d=0, q=0) # AR(9)
sarima.for(diff(log(all_p)), n.ahead=24, p=10, d=0, q=0) # AR(10)
sarima.for(diff(log(all_p)), n.ahead=24, p=11, d=0, q=0) # AR(11)
sarima.for(diff(log(all_p)), n.ahead=24, p=12, d=0, q=0) # AR(12)
```


Checking the Model Conditions: All Poultry

```{r}
sarima(diff(log(all_p)), p=1, d=0, q=0)# AR(1)
sarima(diff(log(all_p)), p=2, d=0, q=0) # AR(2)
sarima(diff(log(all_p)), p=3, d=0, q=0) # AR(3)
sarima(diff(log(all_p)), p=4, d=0, q=0) # AR(4)
sarima(diff(log(all_p)), p=5, d=0, q=0) # AR(5)
sarima(diff(log(all_p)), p=6, d=0, q=0) # AR(6)
sarima(diff(log(all_p)), p=7, d=0, q=0) # AR(7)
sarima(diff(log(all_p)), p=8, d=0, q=0) # AR(8)
sarima(diff(log(all_p)), p=9, d=0, q=0) # AR(9)
sarima(diff(log(all_p)), p=10, d=0, q=0) # AR(10)
sarima(diff(log(all_p)), p=11, d=0, q=0) # AR(11)
sarima(diff(log(all_p)), p=12, d=0, q=0) # AR(12)
```

#### Yule-Walker Estimation

```{r}
# all red meat
rmp.yw = ar.yw(all_rmp, order=12)
rmp.yw$x.mean # mean estimate
rmp.yw$ar # phi parameter estimates
sqrt(diag(rmp.yw$asy.var.coef)) # their standard errors

# all poultry
p.yw = ar.yw(all_p, order=12)
p.yw$x.mean # mean estimate
p.yw$ar # phi parameter estimates
sqrt(diag(p.yw$asy.var.coef)) # their standard errors
```
